Results

### 2022-07-26 article/merge.Rmd
full draft

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
author: 
  - Nicholas Medina ^a^*
  - Max Villalobos ^b^
  - Andrea Johnson ^b^
date: 
  scriptsize\ ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453;
  ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

@@

<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.5), Biotropica (2.5), Eco Evo -->

@@

# Acknowledgements{-}

Thanks to Professor Dan Perlman for initial networking; 
<!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  -->
Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance; 
concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support; 
peer James Fifer for discussion of early draft; 
and Luisa Valdez for translation assistance.

# Author contributions{-}

MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis;
NM wrote initial draft and later revisions.

>>>>

---
title: "Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest"
# author: 
  # - Nicholas Medina * #^a^
  # - Emma Capetz ^c^
  # - Andrea Johnson ^b^
  # - Erin Lau ^c^
  # - Max Villalobos ^b^
# date: \scriptsize
#   ^a^ Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA 02453 USA; 
  # ^b^ Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica
  # ^c^ Earth and Environmental Sciences, Lehigh University, 27 Memorial Drive West, Bethlehem, PA 18015 USA
  # Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI 48109 USA
# author: 
#   - name: "Nicholas Medina"
#     email: nmedina@umich.edu
#     affiliation: Brandeis University
#     orcid_id: 0000-0001-5465-3988
# address:
#   - code: Brandeis University
#     address: Biology and Environmental Studies, 415 South St, Waltham, MA 02453
# keywords: [1, 2, 3]

@@

<!-- journals: J [Appl] Ecol, J [Appl] Veg Sci, Basic Appl Ecol (3.7), Biotropica (2.5), Small-scale Forestry (1.3), Eco Evo -->

@@

<!-- # Acknowledgements{-} -->

<!-- Thanks to Professor Dan Perlman for initial networking;  -->
<!-- <!-- concurrent Osa Conservation managers Max Villalobos and Andrea Johnson for project guidance and supervision;  --> -->
<!-- Lehigh University Professor Don Morris and interns Emma Capetz and Erin Lau along with field station staff Agustín Mendoza for direct field assistance;  -->
<!-- concurrent Osa Conservation general manager Manuel Ramirez and Piro Biological Station staff Juan Carlos Cruz Díaz, Annia Barrantes, Larry Villalobos, Hansel Vargas, and visitors for field support;  -->
<!-- peer James Fifer for discussion of early draft;  -->
<!-- and Luisa Valdez for translation assistance. -->

<!-- # Author contributions{-} -->

<!-- MV and AJ contributed to study conceptualization, administration, and supervision; NM and AJ contributed to funding acquisition; NM contributed to data collection; NM and MV contributed to data analysis; -->
<!-- NM wrote initial draft and later revisions. -->

### 2022-07-26 article/merge.tex
full draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}

@@

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

@@

Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production \emph{(\protect\hyperlink{ref-others}{\textbf{others?}})}, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
Overall, while edge effects near degraded habitat edges are often negative, meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and -community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.

@@

\includegraphics{merge_files/figure-latex/map-1.pdf}

@@

Aboveground biomass did not tend to change with distance to primary forest edge
(Fig \ref{fig:biomass}a), despite significant trends with underlying related variables
(Fig \ref{fig:biomass}).
Median plot wood density tended to increase marginally significantly
(p=0.598)
with edge distance by
\ensuremath{8.2\times 10^{-5}} ±
\ensuremath{1.4\times 10^{-4}} per m
explaining
R2x100\% of variance
among median distance strata values
(Fig \ref{fig:biomass}b).
Plot stem density also tended to increase significantly
(p=0.82)
with edge distance but more strongly and non-linearly, by
\textasciitilde-65.9 ±
78.9 per m
(Fig \ref{fig:biomass}c).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
(Fig \ref{fig:biomass}d).

@@

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/biomass-1.pdf}
\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
\end{figure}

@@

Overall diversity changed significantly with distance to primary forest edge
(Fig \ref{fig:taxa}).
Diversity significantly
(p=0.414)
decreased non-linearly by
\textasciitilde0.1 ±
0.1 per m
as distance to primary forest edge increased,
which explained
R2x100\% of variance
among distance strata medians
(Fig \ref{fig:taxa}a) --
while in contrast taxonomic richness increased slightly
(p=0.067)
and linearly
by
0 ±
0 per m
and had R2x100\% variance explained
among distance strata
(Fig \ref{fig:taxa}b).
Community composition and beta diversity also changed significantly
(p=0.01) with
25.6\%
variance among distance strata explained by distance to primary forest edge,
and the first two principal components explaining
59\% and
24\% totaling
83\%
of variance among calculated plot distances
(Fig \ref{fig:taxa}d).
Key abundant taxa \emph{Vochysia} and \emph{Ficus} showed different responses -- \emph{Vochysia} nearly tended to decrease with distance to primary forest edge
(Fig \ref{fig:taxa}e),
while \emph{Ficus} decreased marginally significantly
(p=0.088)
and linearly by
-62.6 ±
25.1 per m
with R2x100\% variance among distance strata explained
(Fig \ref{fig:taxa}f).

@@

Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest
(Fig \ref{fig:traits}).
Taxa associated with both early and late successional stages decreased significantly
(p=0.042)
and curvi-linearly
by
\textasciitilde{} -584.9 ±
167.3 kg per m
with distance to edge explaining
R2x100\% variance
among distance strata medians
(Fig \ref{fig:traits}a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge
(Fig \ref{fig:traits}b).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
\date{\scriptsize \textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}

@@

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

@@

Properties of the existing tree community also depend on land-use history and previous management \emph{(\protect\hyperlink{ref-omeja12}{Omeja et al., 2012})} including plantation use for timber production, which historically has favored tree taxa that tend to pioneering, or be faster-growing, have lower wood densities, and be less shade-tolerant \emph{(\protect\hyperlink{ref-tabarelli08}{Tabarelli et al., 2008})}.
Legacy effects from management like intensive timber planting could in theory delay or result in non-linear changes between successional stages \emph{(\protect\hyperlink{ref-albrich20}{Albrich et al., 2020}; \protect\hyperlink{ref-gough21}{Gough et al., 2021})}, which would make future management more uncertain, unlike more sustainable thinning practices and agroforestry \emph{(\protect\hyperlink{ref-lefland18}{Lefland et al., 2018})}.
Overall, while edge effects near degraded habitat edges are often negative, meta-population \emph{(\protect\hyperlink{ref-levins79}{Levins, 1979})} and -community \emph{(\protect\hyperlink{ref-leibold04}{Leibold et al., 2004}; \protect\hyperlink{ref-warren15}{Warren et al., 2015})} theories highlight dispersal as a key process for offsetting extinction debt and thus maintaining biodiversity.
Near primary forest, dispersal may facilitate secondary forest regeneration, and help explain relatively high biodiversity compared to adjacent stands that can be found in agroforests \emph{(\protect\hyperlink{ref-oliveira-neto17}{Oliveira-Neto et al., 2017})}, forests affected by logging \emph{(\protect\hyperlink{ref-clark12}{Clark and Covey, 2012}; \protect\hyperlink{ref-edwards14}{Edwards et al., 2014})}, and timber plantations \emph{(\protect\hyperlink{ref-pryde15}{Pryde et al., 2015})}.

@@

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/map-1.pdf}
\caption{\label{fig:map}Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2009 Google, © 2022 Osa Conservation. Photo credit: Nicholas Medina.}
\end{figure}

@@

Aboveground biomass did not tend to change with distance to primary forest edge
(Fig \ref{fig:biomass}a), despite significant trends with underlying related variables
(Fig \ref{fig:biomass}).
Median plot wood density tended to increase marginally significantly
(p=0.105)
with edge distance by
0.00019
per m
explaining
\emph{\textsubscript{10\emph{\% of variance
among median distance strata values
(Fig \ref{fig:biomass}b),
while tissue carbon did not change significantly with distance to forest edge
(Fig \ref{fig:biomass}c).
Plot stem density also tended to increase significantly
(p=0.03)
with edge distance but more strongly and non-linearly, by
\textasciitilde17.4 ±
3.4 per m
(Fig \ref{fig:biomass}d).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
(Fig \ref{fig:biomass}e).
Notably, tree diversity decreased marginally significantly
(p=\textless0.0001)
by
\textasciitilde{} 1
per g plot wood density,
yet only explaining
}}4}\% of variance
(Fig \ref{fig:biomass}f).

@@

\begin{figure}
\centering
\includegraphics{merge_files/figure-latex/biomass-1.pdf}
\caption{\label{fig:biomass} Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians.}
\end{figure}

@@

Overall diversity changed significantly with distance to primary forest edge
(Fig \ref{fig:taxa}).
Diversity significantly
(p=0.003)
decreased non-linearly by
\textasciitilde0.5 ±
0 per m
as distance to primary forest edge increased,
which explained
\emph{73}\% of variance
among distance strata medians
(Fig \ref{fig:taxa}a) --
while in contrast taxonomic richness increased slightly
(p=0.067)
and linearly
by
0 ±
0 per m
and had \emph{16}\% variance explained
among distance strata
(Fig \ref{fig:taxa}b).
Community composition and beta diversity also changed significantly
(p=0.01) with
11.4\%
variance among distance strata explained by distance to primary forest edge,
and the first two principal components explaining
47\% and
23\% totaling
70\%
of variance among calculated plot distances
(Fig \ref{fig:taxa}d).
Key abundant taxa \emph{Vochysia} and \emph{Ficus} showed different responses -- \emph{Vochysia} nearly tended to decrease with distance to primary forest edge
(Fig \ref{fig:taxa}e),
while \emph{Ficus} decreased marginally significantly
(p=0.088)
and linearly by
-62.6 ±
25.1 per m
with \emph{\textasciitilde20}\% variance among distance strata explained
(Fig \ref{fig:taxa}f).

@@

Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest
(Fig \ref{fig:traits}).
Taxa associated with both early and late successional stages decreased significantly
(p=0.042)
and curvi-linearly
by
\textasciitilde{} -584.9 ±
167.3 kg per m
with distance to edge explaining
\emph{\textasciitilde14}\% variance
among distance strata medians
(Fig \ref{fig:traits}a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge
(Fig \ref{fig:traits}b).

### 2022-07-26 article/methods.Rmd
full draft

---
title: "Methods"
output: 
  # html_notebook
  bookdown::html_document2:
    number_sections: FALSE
bibliography: 
- ../../library.bib
- ../../grateful-refs.bib
- ../../software.bib
link-citations: TRUE
---

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. `r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
(See *@taylor15* for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site of (c) Bombacopsis quinata timber plantation on Osa Peninsula, Costa Rica. Map image data © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```
<!-- ![**Fig 1**](../figs/map.png) -->

>>>>

---
title: "Methods"
output: 
  # html_notebook
  # bookdown::html_document2:
  #   number_sections: FALSE
  bookdown::pdf_document2:
    number_sections: FALSE
    toc: FALSE
bibliography: 
- ../../library.bib
- ../../grateful-refs.bib
- ../../software.bib
link-citations: TRUE
---

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. `r RgoogleMaps::getGeoCode("cabo matapalo")`) near the Piro Biological Station run by Osa Conservation.
(See *@taylor15* for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of *Bombacompsis quinata* abandoned in ~1990 after the vulnerably dry-adapted species from the Guanacaste region *[@perezcordero02; @hulshof20]* grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \@ref(fig:map)) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

```{r map, fig.width=4, fig.height=4, fig.cap="Map of (a) conservation areas and (b) study site (c) showing plot design among distance strata, and focal plantation timber taxon (c) Bombacopsis quinata on Osa Peninsula, Costa Rica. Map image data © 2009 Google, © 2022 Osa Conservation. Photo credit: Nicholas Medina."}
source(here::here("figs/map.R"));
sitePlot
# knitr::include_graphics(here::here("figs/figmap.png"))
# magick::image_read(path = "../figs/map.png") %>% plot()
```

<!-- ![**Fig 1**](../figs/map.png) -->

### 2022-07-26 article/results.Rmd
full draft

```{r fig1data}
wd <- filter(plotResultsTbl1, variable == "spg") #alias
stems <- filter(plotResultsTbl1, variable == "stems") #alias
stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
```

Aboveground biomass did not tend to change with distance to primary forest edge 
(Fig \@ref(fig:biomass)a), despite significant trends with underlying related variables 
(Fig \@ref(fig:biomass)).
Median plot wood density tended to increase marginally significantly 
(`r glue("p=",round(as.numeric(wd$pval),3))`) 
with edge distance by 
`r wd$statPrint[[1]]$estimate[[2]] %>% p_round()` ± 
`r wd$statPrint[[1]]$std.error[[2]] %>% rstatix::p_round()` per m 
explaining 
*`r "~10"`*% of variance 
among median distance strata values
(Fig \@ref(fig:biomass)b).
Plot stem density also tended to increase significantly 
(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
with edge distance but more strongly and non-linearly, by 
~`r stems12$statPrint[[1]]$estimate[[2]]` ± 
`r stems12$statPrint[[1]]$std.error[[2]]` per m 
(Fig \@ref(fig:biomass)c).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
(Fig \@ref(fig:biomass)d).

```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) stem density, and (d) canopy light availability. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians  ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
(Fig \@ref(fig:traits)).
Taxa associated with both early and late successional stages decreased significantly 
(`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 
and curvi-linearly 
by 
~ `r succession$statPrint[[7]]$estimate[[2]]` ± 
`r succession$statPrint[[7]]$std.error[[2]]` kg per m 
with distance to edge explaining 
`r "R2x100"`% variance 
among distance strata medians 
(Fig \@ref(fig:traits)a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
(Fig \@ref(fig:traits)b).

>>>>

```{r fig1data}
wd <- filter(plotResultsTbl1, variable == "spg") #alias
# stems <- filter(plotResultsTbl1, variable == "stems") #alias
stems12 <- filter(plotResultsTbl12, variable == "stems") %>% distinct() #alias
divwd <- filter(covarResultsTbl1, variable == "entropy")
```

Aboveground biomass did not tend to change with distance to primary forest edge 
(Fig \@ref(fig:biomass)a), despite significant trends with underlying related variables 
(Fig \@ref(fig:biomass)).
Median plot wood density tended to increase marginally significantly 
(`r glue("p=",round(as.numeric(wd$pval),3))`) 
with edge distance by 
`r wd$statPrint[[1]]$estimate[[2]] %>% rstatix::p_format()` per m 
<!-- ± `r wd$statPrint[[1]]$std.error[[2]] |> round(3) #formatting...` -->
explaining 
*`r "~10"`*% of variance 
among median distance strata values
(Fig \@ref(fig:biomass)b),
while tissue carbon did not change significantly with distance to forest edge 
(Fig \@ref(fig:biomass)c).
Plot stem density also tended to increase significantly 
(`r glue("p=",round(as.numeric(stems12$pval[[2]][2]),2))`) 
with edge distance but more strongly and non-linearly, by 
~`r stems12$statPrint[[1]]$estimate[[2]]` ± 
`r stems12$statPrint[[1]]$std.error[[2]]` per m 
(Fig \@ref(fig:biomass)d).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge 
(Fig \@ref(fig:biomass)e).
Notably, tree diversity decreased marginally significantly 
(`r glue("p=", divwd$pval[[1]] |> rstatix::p_format())`)
by 
~ `r divwd$statPrint[[1]]$estimate[[2]] |> rstatix::p_round()` 
<!-- ± `r divwd$statPrint[[1]]$std.error[[2]] |> rstatix::p_round()` -->
per g plot wood density, 
yet only explaining
*`r "~4"`*% of variance
(Fig \@ref(fig:biomass)f).

```{r biomass, fig.width=3.5, fig.height=3.5, fig.cap=" Plot stand properties, namely (a) biomass, (b) wood density, (c) tissue carbon, (d) stem density, (d) canopy light availability, all across distance to primary forest edge gradient, and (e) diversity against wood density. Grey dots show individual tree values, white dots with border show plot medians, black dots show distance strata medians ± 1 absolute deviation, and black lines show linear or non-linear regression through strata medians."}
source(here::here("figs/fig1.R"), echo = F); 
fig1
# knitr::include_graphics(here::here("figs/fig1.png"))
```

@@

Overall trait regeneration highlighted successional stage associations over primary dispersal mode along distance to primary forest 
(Fig \@ref(fig:traits)).
Taxa associated with both early and late successional stages decreased significantly 
(`r glue("p=",round(as.numeric(succession$pval[[7]][2]),3))`) 
and curvi-linearly 
by 
~ `r succession$statPrint[[7]]$estimate[[2]]` ± 
`r succession$statPrint[[7]]$std.error[[2]]` kg per m 
with distance to edge explaining 
*`r "~14"`*% variance 
among distance strata medians 
(Fig \@ref(fig:traits)a).
Dispersal modes did not show consistent trends in biomass with increasing distance to edge 
(Fig \@ref(fig:traits)b).
